---
title: "MSE 246 Data Summary"
author: "Samuel Hansen"
date: "1/16/2017"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, cache = TRUE)
```

```{r}
# Initialize libraries and input files 
library(ggrepel)
library(knitr)
library(lubridate)
library(tidyverse)
file_in <- "../data/SBA_Loan_data_1.csv"
```

```{r}
# Read in data 
df <- 
  read_csv(file_in) %>%
  plyr::rename(replace = c("2DigitNAICS" = "NAICS")) %>%
  mutate(ApprovalDate = mdy(ApprovalDate))
```

#Fraction of Defaulted Loans 
```{r}
kable(
  df %>%
  count(LoanStatus) %>%
  mutate(proportion = n/sum(n))
)
```

#NAICS Code 

##Default Rate by NAICS Code 
```{r, fig.width=10, fig.height=6}
df %>%
  group_by(NAICS, year = year(ApprovalDate)) %>%
  summarise(default_rate = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = year, y = default_rate, color = NAICS)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Approval Year", y = "Default Rate", title = "Default Rate by NAICS Code")
```

#Loan Amount

##Loan Amount Histogram
```{r}
df %>%
  ggplot(mapping = aes(x = GrossApproval)) +
  geom_histogram(binwidth = 10000) +
  scale_x_continuous(labels = scales::dollar, limits = c(0,2000000)) +
  labs(x = "Gross Approval Amount", y = "Count",
       title = "Loan Gross Approval Histogram")
```

##Default Rate by Loan Amount
```{r}
loan_labels <- c("(-10,1e+05]" = "< 100k",
                 "(1e+05,3e+05]" = "100k - 300k",
                 "(3e+05,5e+05]" = "300k - 500k",
                 "(5e+05,1e+06]" = "500k - 1m",
                 "(1e+06,2e+06]" = "1m - 2m",
                 "(2e+06,4e+06]" = "> 2m")
df %>%
  mutate(#loan_bin = cut_width(GrossApproval, width = 1000000),
         loan_bin = cut(GrossApproval, breaks = c(-10, 100000, 300000, 
                                                  500000, 1000000,2000000, 4000000)),
         loan_bin = plyr::revalue(loan_bin, loan_labels),
         year = year(ApprovalDate)) %>%
  
  group_by(loan_bin, year) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = year, y = frac_defaulted, color = loan_bin)) +
  geom_point(mapping = aes(size = count)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(name = "Number of Loans") + 
  scale_color_discrete(name = "Loan Amount ($)") + 
  labs(x = "Approval Year", y = "Default Rate", title = "Default Rate by Loan Amount")
```

#Subprogram Description

##Default Rate vs. Subprogram Description

Subprogram description is the specific subprogram that the loan was approved 
under. See SOP 50 10 5 for definitions and rules for each subprogram.

```{r}
df %>%
  group_by(subpgmdesc) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = count, y = frac_defaulted, color = subpgmdesc)) +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::comma) +
  scale_color_discrete(name = "Subprogram Type") + 
  labs(x = "Count of Subprogram", y = "Default Rate",
       title = "Default Rate vs. Subprogram Type")
```

This plot suggests we can collapse the `PSF` and `Refinance` categories into `other` because they have low counts and default rates. 

#Business Type

##Default Rate vs. Business Type

```{r}
df %>%
  group_by(BusinessType) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = count, y = frac_defaulted, color = BusinessType)) +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::comma) +
  scale_color_discrete(name = "Business Type") + 
  labs(x = "Count of Business Type", y = "Default Rate",
       title = "Default Rate vs. Business Type")
```

##Default Rate vs. Business Type by Approval Year

```{r, fig.width=10, fig.height=6}
df %>%
  group_by(BusinessType, year = year(ApprovalDate)) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = year, y = frac_defaulted, color = BusinessType)) +
  geom_point(mapping = aes(size = count)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(name = "Count") +
  # scale_x_continuous(labels = scales::comma) +
  scale_color_discrete(name = "Business Type") + 
  labs(x = "Approval Year", y = "Default Rate",
       title = "Default Rate vs. Business Type by Approval Year")
```

#Loan Term

##Load Term Histogram

```{r}
df %>%
  ggplot(mapping = aes(x = TermInMonths)) +
  geom_histogram(binwidth = 12) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0,400,24)) +
  labs(x = "Loan Term in Months", y = "Count",
       title = "Loan Term Histogram")
```

##Default Rate vs. Load Term by Approval Year

```{r}
df %>%
  mutate(term_bin = cut_width(TermInMonths, width = 48), 
         term_bin = plyr::revalue(term_bin, 
                                  replace = c("[-24,24]" = "[0,24]"))) %>%
  group_by(term_bin) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = count, y = frac_defaulted, color = term_bin)) +
  geom_point() +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_discrete(name = "Loan Term (Months)") + 
  labs(x = "Count of Term Bin", y = "Default Rate",
       title = "Default Rate vs. Loan Term")
```

The bin (120,168] has an unusually high default rate; however, it has a low
count. In fact, all term bins except for (216,264] and (72,120] have less than
100 observations. In turn, we may consider collapsing these infrequent bins,
rather than treating `Term in Months` as a continuous feature. 

#Matching Lending State

##Default Rate vs. Matching Lending State by Approval Year

```{r, fig.width=10, fig.height=6}
df %>%
  group_by(SameLendingState, year = year(ApprovalDate)) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = year, y = frac_defaulted, color = SameLendingState)) +
  geom_point(mapping = aes(size = count)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(name = "Count") + 
  labs(x = "Approval Year", y = "Default Rate", 
       title = "Default Rate by Lending State Match and Approval Year")
```

#Multi-Time Borrower

##Default Rate vs. Multi-Time Borrower Status by Approval Year

```{r, fig.width=10, fig.height=6}
df %>%
  group_by(MultiTimeBorrower, year = year(ApprovalDate)) %>%
  summarise(count = n(),
            frac_defaulted = mean(LoanStatus == "CHGOFF")) %>%
  ggplot(mapping = aes(x = year, y = frac_defaulted, color = MultiTimeBorrower)) +
  geom_point(mapping = aes(size = count)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(name = "Count") + 
  labs(x = "Approval Year", y = "Default Rate", 
       title = "Default Rate by Multi-Time Borrower Status and Approval Year")
```
