---
title: "MS&E 246 Report"
output: html_document
---
# Tranches: Distribution of Losses
In this section, we will estimate the distribution for the one and five year losses of an investor who has purchased a $[5\%, 15\%]$ tranche backed by the 500 loan portfolio.  In addition, we will investigate the loss distribution of the $[15\%, 100\%]$ senior tranche. 

### Portfolio and assumptions
We assume that all active loans whose term length does not expire within the 1- and/or 5-year window are eligible for the tranche.  We select from the dataframe of total loans, a subset of active loans that meet this requirement.  

```{r eval = FALSE}
# Clear our workspace
graphics.off()
rm(list = ls())

# Read in our data files
df = readRDS('tranche_prob_df.rds')
df2 = readRDS('merged.rds')

# Create new data frame [Approval Year, Approval Amount, Default Year (NA if no def), Termination Year]
mat = rep(0,0); # Initialize
mat = cbind(mat, as.numeric(levels(df[,3]))[df[,3]]); # Append approval year
mat = cbind(mat, df[,4]); # Append approval amt
mat = cbind(mat, as.numeric(format(df2$ChargeOffDate, '%Y'))); # Append Default Year
mat = cbind(mat, as.numeric(levels(df[,3]))[df[,3]] + ceiling(df2$TermInMonths/12)) # Append term year

# Determine active loans for given year 
year = 1990
# Initialize matrices for active loans
active1 = rep(0,0);
active5 = rep(0,0);
# Loop over entire dataframe of loans
for (i in 1:54794){
  # If approval year <= test year AND it did not default prior to test year
  if ((mat[i,1] <= year) && (is.na(mat[i,3]) || mat[i,3]>= year)){
    # If the loan has not expired within a year
    if (mat[i,4] >= year+1){
      # Append loan to active loans for 1-year
      active1 = rbind(active1, mat[i,])
    }
    # If the loan has not expired within 5 years
    if (mat[i,4] >= year+5){
      # Append loan to active loans for 5-year
      active5 = rbind(active5, mat[i,])
    }
  }
}
```

#### Selection of loans for portfolio
Once the dataframe of eligible loans for the portfolio has been created, we select 500 loans uniformly random from the list.  We store the 500 loans in a matrix in R.  

```{r eval = FALSE}
# Pick 500 loans uniformly random from active loan list
sim1 = sample(1:dim(active1)[1], 500)
sim1 = active1[sim1,]
sim5 = sample(1:dim(active5)[1], 500)
sim5 = active5[sim5,]
```

#### Determine value of the portfolio of loans 
Once we have our loans selected for our portfolio, we determine the value of the portfolio.  It may seem intuitive to simply add the value of each loan for the portfolio to determine the value of the tranche.  However, this method would not account for different term lengths.  For example, a loan for \$100,000 over 1-year would be more be more valuable in the 1-year tranche than a 5-year loan for \$200,000.  We account for this problem by normalizing the value of each loan by the term length.  Note, this will ignore minor discrepancies between accrued interest.  In addition, we assume that loans either default or are paid in full at the loan termination date.  Note, this assumption ignores the possibility of a borrower paying the loan off before the loan due date.

```{r eval = FALSE}
# Determine total value of portfolio
value1 = sum(sim1[,2]/(sim1[,4]-sim1[,1]))
value5 = sum(sim5[,2]/(sim5[,4]-sim5[,1]))
```

#### Determine the loss from the portfolio of loans
In an indentical manner to determining the value of the portfolio, we will determine the loss observed by the portfolio.  

```{r eval = FALSE}
# Determine default loans 
def1 = sim1[!is.na(sim1[,3]),]
def5 = sim5[!is.na(sim5[,3]),]

# Determine total loss
loss1 = sum(def1[,2]/(def1[,4]-def1[,1]))
loss5 = 0;
if (is.null(dim(def5))){
  loss5 = def5[2]/(def5[4]-def[1])
} else {
  if ((dim(def5)[1]) > 1){
    for (i in 1:(dim(def5)[1])){
      loss5 = loss5 + def5[i,2]/(def5[i,4]-def5[i,1])
    }
  }
}
# Determine absolute ratio of loss
loss_percent1 = c(loss_percent1, loss1/value1);
loss_percent5 = c(loss_percent5, loss5/value5);
```

#### Generate loss distribution and plotting
We run this simulation of selected 500 loans uniformly random from the list of active loans 1000 times, and compute the appropriate losses for each tranche.  We then plot the approximated distribution using the Kernel Density Estimator (KDE).
```{r eval = FALSE}
# Initialize Vector of Tranche Losses
tranche1_5_15 = rep(0, length(loss_percent1))
tranche1_15_100 = rep(0, length(loss_percent1))
tranche5_5_15 = rep(0, length(loss_percent1))
tranche5_15_100 = rep(0, length(loss_percent5))

# Tranche [.05, .15] loss for 1 year
for (i in 1:length(loss_percent1)){
  if (loss_percent1[i]<.05){
    tranche1_5_15[i] = 0;
  } else {
    if (loss_percent1[i] > 0.15){
      tranche1_5_15[i] = 1;
    } else{
      tranche1_5_15[i] = 10*(loss_percent1[i] - .05)
    }
  }
}

# Tranche [.05, .15] loss for 5 year
for (i in 1:length(loss_percent5)){
  if (loss_percent5[i]<.05){
    tranche5_5_15[i] = 0;
  } else {
    if (loss_percent5[i] > 0.15){
      tranche5_5_15[i] = 1;
    } else{
      tranche5_5_15[i] = 10*(loss_percent5[i] - .05)
    }
  }
}

# Tranche [.15, 1] loss for 1 year
for (i in 1:length(loss_percent1)){
  if (loss_percent1[i]<.15){
    tranche1_15_100[i] = 0;
  } else {
    tranche1_15_100[i] = 100/85*(loss_percent1[i] - .15)
  }
}

# Tranche [.15, 1] loss for 1 year
for (i in 1:length(loss_percent5)){
  if (loss_percent5[i]<.15){
    tranche5_15_100[i] = 0;
  } else {
    tranche5_15_100[i] = 100/85*(loss_percent5[i] - .15)
  }
}

# Plot results and save figures
y_max = max(c(max(density(tranche1_5_15, bw = .0075, from = 0, to = 1)$y), max(density(tranche1_15_100, bw = .0075, from = 0, to = 1)$y)))
png(filename = paste("./DefaultDist/5yr", year, "tranche.png", sep = "_"))
plot(density(tranche1_5_15, bw = .0075, from = 0, to = 1), 
     main = paste("Loss for 1-year Portfolio in", year, "for tranches [5,15] and [15,100]", sep = " "), 
     xlab = "Loss", ylab = "Density", 
     xlim = c(0,1), ylim = c(0, min(c( 15, y_max))), col = 'blue');
lines(density(tranche1_15_100, bw = .0075, from = 0, to = 1), col = 'red')
legend('top', c('Tranche [0,15]', 'Tranche [15,100]'), col = c('blue', 'red'), lwd = 2)
dev.off()

y_max = max(c(max(density(tranche5_5_15, bw = .0075, from = 0, to = 1)$y), max(density(tranche5_15_100, bw = .0075, from = 0, to = 1)$y)))
png(filename = paste("./DefaultDist/1yr", year, "tranche.png", sep = "_"))
plot(density(tranche5_5_15, bw = .0075, from = 0, to = 1), 
     main = paste("Loss for 5-year Portfolio in", year, "for tranches [5,15] and [15,100]", sep = " "), 
     xlab = "Loss", ylab = "Density", 
     xlim = c(0,1), ylim = c(0, min(c( 15, y_max))), col = 'blue');
lines(density(tranche5_15_100, bw = .0075, from = 0, to = 1), col = 'red')
legend('top', c('Tranche [5,15]', 'Tranche [15,100]'), col = c('blue', 'red'), lwd = 2)
dev.off()
```

### Complete R-Code with simulation loops
```{r eval = TRUE}
graphics.off()
rm(list = ls())


df = readRDS('tranche_prob_df.rds')
df2 = readRDS('merged.rds')

# Create new data frame [Approval Year, Approval Amount, Default Year (NA if no def), Termination Year]
mat = rep(0,0); # Initialize
mat = cbind(mat, as.numeric(levels(df[,3]))[df[,3]]); # Append approval year
mat = cbind(mat, df[,4]); # Append approval amt
mat = cbind(mat, as.numeric(format(df2$ChargeOffDate, '%Y'))); # Append Default Year
mat = cbind(mat, as.numeric(levels(df[,3]))[df[,3]] + ceiling(df2$TermInMonths/12)) # Append term year

# Generate a simulation
# Determine active loans
cat("\014")
#for (year in 1990:2013){
for (year in 2005){
  #print(year)
  active1 = rep(0,0);
  active5 = rep(0,0);
  for (i in 1:54794){
    if ((mat[i,1]<=year) && (is.na(mat[i,3]) || mat[i,3]>= year)){
      if (mat[i,4] >= year+1){
        active1 = rbind(active1, mat[i,])
      }
      if (mat[i,4] >= year+5){
        active5 = rbind(active5, mat[i,])
      }
    }
  }
  
  loss_percent1 = rep(0,0)
  loss_percent5 = rep(0,0)
  for (iter in 1:10000){
    
    # Pick 500 loans uniformly random from active loan list
    sim1 = sample(1:dim(active1)[1], 500)
    sim1 = active1[sim1,]
    sim5 = sample(1:dim(active5)[1], 500)
    sim5 = active5[sim5,]
    
    # Determine total value of portfolio
    value1 = sum(sim1[,2]/(sim1[,4]-sim1[,1]))
    value5 = sum(sim5[,2]/(sim5[,4]-sim5[,1]))
    
    # Determine default loans 
    def1 = sim1[!is.na(sim1[,3]),]
    def5 = sim5[!is.na(sim5[,3]),]
    
    # Determine total loss
    loss1 = sum(def1[,2]/(def1[,4]-def1[,1]))
    loss5 = 0;
    if (is.null(dim(def5))){
      loss5 = def5[2]/(def5[4]-def[1])
    } else {
      if ((dim(def5)[1]) > 1){
        for (i in 1:(dim(def5)[1])){
          loss5 = loss5 + def5[i,2]/(def5[i,4]-def5[i,1])
        }
      }
    }
    # Determine ratio of loss
    loss_percent1 = c(loss_percent1, loss1/value1);
    loss_percent5 = c(loss_percent5, loss5/value5);
  }
  
  # Initialize Vector of Tranche Losses
  tranche1_5_15 = rep(0, length(loss_percent1))
  tranche1_15_100 = rep(0, length(loss_percent1))
  tranche5_5_15 = rep(0, length(loss_percent1))
  tranche5_15_100 = rep(0, length(loss_percent5))
  
  # Tranche [.05, .15] loss for 1 year
  for (i in 1:length(loss_percent1)){
    if (loss_percent1[i]<.05){
      tranche1_5_15[i] = 0;
    } else {
      if (loss_percent1[i] > 0.15){
        tranche1_5_15[i] = 1;
      } else{
        tranche1_5_15[i] = 10*(loss_percent1[i] - .05)
      }
    }
  }
  
  # Tranche [.05, .15] loss for 5 year
  for (i in 1:length(loss_percent5)){
    if (loss_percent5[i]<.05){
      tranche5_5_15[i] = 0;
    } else {
      if (loss_percent5[i] > 0.15){
        tranche5_5_15[i] = 1;
      } else{
        tranche5_5_15[i] = 10*(loss_percent5[i] - .05)
      }
    }
  }
  
  # Tranche [.15, 1] loss for 1 year
  for (i in 1:length(loss_percent1)){
    if (loss_percent1[i]<.15){
      tranche1_15_100[i] = 0;
    } else {
      tranche1_15_100[i] = 100/85*(loss_percent1[i] - .15)
    }
  }
  
  # Tranche [.15, 1] loss for 1 year
  for (i in 1:length(loss_percent5)){
    if (loss_percent5[i]<.15){
      tranche5_15_100[i] = 0;
    } else {
      tranche5_15_100[i] = 100/85*(loss_percent5[i] - .15)
    }
  }
}

```
#### Plot results
```{r}
# Plot results and save figures
y_max = max(c(max(density(tranche1_5_15, bw = .0075, from = 0, to = 1)$y), max(density(tranche1_15_100, bw = .0075, from = 0, to = 1)$y)))
#png(filename = paste("./DefaultDist/1yr", year, "tranche.png", sep = "_"))
plot(density(tranche1_5_15, bw = .0075, from = 0, to = 1), 
     main = paste("Loss for 1-year Portfolio in", year, "for tranches [5,15] and [15,100]", sep = " "), 
     xlab = "Loss", ylab = "Density", 
     xlim = c(0,1), ylim = c(0, min(c( 15, y_max))), col = 'blue');
lines(density(tranche1_15_100, bw = .0075, from = 0, to = 1), col = 'red')
legend('top', c('Tranche [0,15]', 'Tranche [15,100]'), col = c('blue', 'red'), lwd = 2)
#dev.off()
```

```{r}
y_max = max(c(max(density(tranche5_5_15, bw = .0075, from = 0, to = 1)$y), max(density(tranche5_15_100, bw = .0075, from = 0, to = 1)$y)))
#png(filename = paste("./DefaultDist/5yr", year, "tranche.png", sep = "_"))
plot(density(tranche5_5_15, bw = .0075, from = 0, to = 1), 
     main = paste("Loss for 5-year Portfolio in", year, "for tranches [5,15] and [15,100]", sep = " "), 
     xlab = "Loss", ylab = "Density", 
     xlim = c(0,1), ylim = c(0, min(c( 15, y_max))), col = 'blue');
lines(density(tranche5_15_100, bw = .0075, from = 0, to = 1), col = 'red')
legend('top', c('Tranche [5,15]', 'Tranche [15,100]'), col = c('blue', 'red'), lwd = 2)
#dev.off()
```


```{r}
y_max = max(c(max(density(tranche1_5_15, bw = .0075, from = 0, to = 1)$y), max(density(tranche5_15_100, bw = .0075, from = 0, to = 1)$y)))
#png(filename = paste("./DefaultDist/full", year, "tranche.png", sep = "_"))
plot(density(tranche1_5_15, bw = .0075, from = 0, to = 1), 
     main = paste("Loss Distribution for 1- and 5-year in", year, "for tranches [5,15], [15,100]", sep = " "), 
     xlab = "Loss", ylab = "Density", 
     xlim = c(0,1), ylim = c(0, min(c( 15, y_max))), col = 'blue');
lines(density(tranche5_5_15, bw = .0075, from = 0, to = 1), col = 'red')
lines(density(tranche1_15_100, bw = .0075, from = 0, to = 1), col = 'green')
lines(density(tranche5_15_100, bw = .0075, from = 0, to = 1), col = 'orange')
legend('top', c('1-yr Tranche [5,15]', '5-yr Tranche [5,15]', '1-yr Tranche [15,100]', '5-yr Tranche [15,100]'), col = c('blue', 'red', 'green', 'orange'), lwd = 2)
#dev.off()
```

### Interpretations and Comparison of Distributions
We can see from the approximated density plots that in the early- to mid-90's, the [5%, 15%] tranche was only slightly more risky than the [15%, 100%] tranche.  Almost half the randomly generated portfolios generated a 0% loss in the [5%, 15%] tranche in 1997.  The senior tranche was statistically loss-less until 2005.  However, by 2007, the [5%, 15%] tranche receives almost 100% loss, and the senior tranche received an average of 10% loss.  From a risk management point of view, one would clearly want to invest in the senior tranche prior to 2005-06.  However, after the financial crisis, the [5%, 15%] tranche almost certainly received 100% loss, and the senior tranche received an average of 10% loss.  
      